
# Some In/Out examples

This chapter gives some examples of how to pass values from R to cpp and vice versa. There is a high degree of flexibility.

## Pass a list

A list can contain different classes of objects. here we want to pass the following list to a C++ function which does nothing else but returning all the objects with a slightly different name in a newly created list.

```{r,tidy=FALSE}
mylist = list(A = array(1:10,c(10,1)),
			  a = pi,
			  b = FALSE,
			  C = matrix(rnorm(20),4,5))
```

You can see that have 4 different classes here: a vector, a double, a boolean and a matrix. Let's pass it to C++:


```{r,tidy=FALSE,message=FALSE}
library(Rcpp)
library(inline)

src2 <- '
	List l(inlist);
	NumericVector one = as<NumericVector>(l["A"]);
	double two = as<double>(l["a"]);
	bool three = as<bool>(l["b"]);
	NumericMatrix four = as<NumericMatrix>(l["C"]);
	// do something useful here
	// ...
	// put together into a list
	// notice: if not using the inline package, must declare
	// namespace Rcpp, or prepend all Rcpp functions with
	// Rcpp::
	List outlist = List::create( _["A.out"] = one, _["a.out"] = two, _["b.out"] = three, _["C.out"] = four);
	return outlist;
	'

f3 <- cxxfunction(signature(inlist="List"),body=src2,plugin="Rcpp")

f3(inlist=mylist)
```
